---
title: "Onassis: Ontology Annotation and Semantic Similarity software"
author: "Eugenia Galeota"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
---

```{r loading_libs}
library(Onassis)
library(GEOmetadb)
library(tools)
```



```{r loading_cistrome_metadata}
setwd("/Users/egaleota/Library/Mobile Documents/com~apple~CloudDocs/IITDrive/OneDriveFondazioneIstitutoItalianoTecnologia/ChromStates")
cistrome_data= read.table('histone_human_data_information.txt', sep='\t', header=F, stringsAsFactor=F)

gsm_ids <- unique(cistrome_data$V2)

geo_con <- connectToGEODB(download=TRUE, destdir='./data')

gsm_metadata <- dbGetQuery(geo_con, paste0("select * from gsm where gsm in ('", paste(gsm_ids, collapse="','"), "')"))

```

To annotate cistrome data with terms from cell line ontology we downloaded the cell line ontology in OWL format from https://raw.githubusercontent.com/CLO-ontology/CLO/master/src/ontology/clo_merged.owl

With the owl ontology available we can create a dictionary for conceptmapper

```{r dictionary_of_cells }

cell_dict <- dictionary(inputFileOrDb = file.path(getwd(), 'data', 'cl.obo'), dictType = 'OBO',outputdir = file.path(getwd(), 'data'), synonymType = 'ALL' )

cell_dict <- dictionary(inputFileOrDb = file.path(getwd(), 'data', 'cmDict-clo_merged.xml'), dictType = 'CMDICT')


efo_dict <- dictionary(inputFileOrDb = './data/efo.obo', dictType = 'OBO', outputdir = './data/', synonymType = 'ALL')

```

The code produced a dictionary cmDict-clo_merged.xml in the data directory.
Now we have to annotate the data frame containing the metadata of the samples available in cistrome 

```{r cell_line_annotation}

opt <- CMoptions()
paramValueIndex(opt) <- 31

#Adding cell to the source name to find also cell names
gsm_metadata$source_name_ch1[!grepl('cell', tolower(gsm_metadata$source_name_ch1))] <- paste0(gsm_metadata$source_name_ch1[!grepl('cell', tolower(gsm_metadata$source_name_ch1))], ' cell')

gsm_metadata$characteristics_ch1 <- gsub('\\t' , ' ', gsm_metadata$characteristics_ch1 )
gsm_metadata$characteristics_ch1 <- gsub(';' , ' ', gsm_metadata$characteristics_ch1 )

#Annotation of cell lines
cell_line_annotations <- annotate(inputFileorDf = gsm_metadata[, c('gsm', 'source_name_ch1', 'characteristics_ch1'),  ], options = opt, dictionary = cell_dict, outDir = file.path(getwd(), 'data', 'tissue'))

cell_line_filtered=cell_line_annotations 
cell_line_filtered <- cell_line_filtered[grep("UBERON|CL", cell_line_filtered$term_id),]
cell_line_filtered <- cell_line_filtered[which(!cell_line_filtered$term_name %in% c('cell', 'tissue', 'Homo sapiens', 'molecule', 'female organism', 'male organism', 'protein', 'cell line cell',  'chromatin', 'signaling', 'cultured cell', 'multicellular organism', 'compound organ', 'organ', 'anatomical conduit')),]
 cell_line_filtered <- cell_line_filtered[!duplicated(cbind(cell_line_filtered$sample_id, cell_line_filtered$term_id)),]


efo_annotations <- annotate(inputFileorDf = gsm_metadata[, c('gsm', 'source_name_ch1', 'characteristics_ch1') ], options = opt, dictionary = efo_dict, outDir = './data/')



#Annotation of diseases 
disease_dict <- dictionary(inputFileOrDb = file.path(getwd(), 'data' , 'doid-non-classified.obo'), dictType = 'OBO',outputdir = file.path(getwd(), 'data'), synonymType = 'ALL' )




disease_annotations <- annotate(inputFileorDf = gsm_metadata[, c('gsm', 'source_name_ch1', 'characteristics_ch1') ], options = opt, dictionary = disease_dict, outDir = file.path(getwd(), 'data'))

healthy_samples <- gsm_metadata$gsm[grep('disease: none|disease: normal|disease: presumed normal|disease: no ad present|disease: no ad evident', tolower(gsm_metadata$characteristics_ch1))]

gsm_metadata2 <- gsm_metadata[which(!gsm_metadata$gsm %in% healthy_samples),]
disease_annotations$term_url[which(disease_annotations$sample_id %in% healthy_samples)] <- 'Healthy'
disease_annotations$term_id[which(disease_annotations$sample_id %in% healthy_samples)] <- 'Healthy'
disease_annotations$term_name[which(disease_annotations$sample_id %in% healthy_samples)] <- 'Healthy'

disease_annotations$matched_sentence[which(disease_annotations$sample_id %in% healthy_samples)] <- ""

disease_annotations <- unique(disease_annotations)

disease_annotations <- disease_annotations[which(!disease_annotations$term_name=='disease'),]
 disease_annotations <- disease_annotations[!duplicated(cbind(disease_annotations$sample_id, disease_annotations$term_id)),]



cell_dt <- setDT(cell_line_filtered)
cell_collapsed <- cell_dt[, lapply(.SD, function(x) toString(unique(x[order(x)]))), by= sample_id]
disease_dt <- setDT(disease_annotations)
disease_collapsed <- disease_dt[, lapply(.SD, function(x) toString(unique(x[order(x)]))), by= sample_id]

saveRDS(cell_collapsed, file='./data/cell_line_collapsed.rds')

saveRDS(disease_collapsed, file='./data/disease_collapsed.rds')



colnames(cell_collapsed)[2] <-"cell_term_id"
colnames(cell_collapsed)[3] <-"cell_term_name"
colnames(cell_collapsed)[4] <- "cell_term_url"
colnames(cell_collapsed)[5] <- "cell_matched_sentence"


total_annotations <- merge(cell_collapsed, disease_collapsed, by='sample_id', all.x=TRUE, all.y=TRUE)

total_annotations_collapsed <- total_annotations[, lapply(.SD, function(x) toString(unique(x[order(x)]))), by= sample_id]

```

```{r similarity}

annotated_samples <- as.character(as.vector(unique(cell_collapsed$cell_term_url)))
n <- length(annotated_samples)
names <- paste0('sample_', seq(1, n, by=1))
 
annotated_df <- as.data.frame(cbind(names, annotated_samples))

ontologyfile <- './data/cl.obo'
samples_results <- data.frame(sample1 = character(0), sample2= character(0), value = double(0L))
samples_results <- matrix(0, nrow=n, ncol=n)
rownames(samples_results) <- colnames(samples_results) <- annotated_samples

n <- length(unique(annotated_df[,2]))

for(i in 1:(n-1)){
  print(paste0('VALORE DI I: ', i))
  sample1 <- as.character(annotated_df[i, 2])
  sample1 <- strsplit(sample1, ', ')[[1]]
  j = i + 1 
  for(k in j:n){
    sample2 <- as.character(annotated_df[k, 2])
    sample2 <- strsplit(sample2, ', ')[[1]]
    two_samples_similarity <- similarity(ontologyfile, sample1, sample2)
    samples_results[i, k] <- samples_results[k, i] <- two_samples_similarity
  }
}
diag(samples_results) <- 1

sample_name_results <- samples_results
colnames(sample_name_results) <- rownames(sample_name_results) <- cell_collapsed[match(rownames(samples_results), cell_collapsed$cell_term_url)]$cell_term_name

sample_name_results <- 1 - sample_name_results
similarity_dist <- as.dist(sample_name_results)
saveRDS(samples_results, file='./data/similarity.rds')
pdf('./data/semantic_distance.pdf',width=80, height=40)
op <- par(mar = c(12,12,12,12) + 0.1)
plot(hclust(d=similarity_dist), main='Semantic similarity distance based clustering of semantic classes associated to Cistrome data', hang=-1, col.axis='#F38630', col.lab="#7c8071", lwd=3, sub='')
dev.off()


semantic_distance <- as.dist(sample_name_results)
clusters <- hclust(d=semantic_distance)
merged_clusters <- cutree(clusters, h=quantile(clusters$height, 0.3))
merged_clusters <- as.data.frame(cbind(names(merged_clusters), merged_clusters))
colnames(merged_clusters) <- c('annotation', 'cluster')
merged_clusters_names <- aggregate( annotation ~ cluster , data=merged_clusters, paste, collapse=', ')
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
nuovi_ids <- apply(merged_clusters_names, 1, function(row) {
  values <- gsub(" ", "", row[2])
  values <- strsplit(values, ',')[[1]]
  values <- unique(values)
  paste(values, collapse=',')
})
merged_clusters_names <- cbind(merged_clusters_names, nuovi_ids)
merged_clusters_names <- unique(merged_clusters_names[, c('cluster', 'nuovi_ids')])
merged_clusters <- merge(merged_clusters, merged_clusters_names, by='cluster', all.x=TRUE)





heatmap.2(samples_results, density.info = "none", trace="none", main='Semantic similarity of annotated samples', margins=c(5,5))



topMarks= names(sort(table(data[,7]), decreasing=TRUE)[1:6])
condNumb= function(targetMarks= topMarks, conds, marks=data[,7]) {
	condsT= sort(table(conds), decreasing=T)
	marksN= rep(0, length(condsT))
	names(marksN)= names(condsT)
	targetMarksN= length(targetMarks)
	for(i in 1:length(condsT)) {
		cond= names(condsT)[i]
		if(cond=='None' || cond=='None # None') next
		if(condsT[i]<targetMarksN) break
		inds= which(conds==cond)
		marksN[i]= length(which(targetMarks %in% marks[inds]))
	}
	condsSel= names(marksN)[which(marksN>=targetMarksN)]
	return(condsSel)
}

pdf('chromStates_conds.pdf', 16,11)
	layout(matrix(1:6,3,2))
	par(mar=c(13,5,3,1))
	cols= c(rep('grey95',4), 'grey80','grey60', rep('grey30', 14))
	sData= sort(table(data[,7]), decreasing=T)
	mp= barplot(sData[1:20], las=3, ylab='#samples', col=cols)
	text(x=mp[4:6], y=sData[4:6]+60, labels=c('top4','top5','top6'))
	for(i in 4:6) {
		conds= data[,i]
		condN6= condNumb(targetMarks= topMarks, conds= conds, marks=data[,7])
		condN5= condNumb(targetMarks= topMarks[1:5], conds= conds, marks=data[,7])
		condN4= condNumb(targetMarks= topMarks[1:4], conds= conds, marks=data[,7])
		condN= paste0('tot=', length(unique(conds)), ' top6=', length(condN6), 
			' top5=', length(condN5), ' top4=', length(condN4))
		barplot(sort(table(conds[conds %in% condN6]), decreasing=T), las=3, main= condN, ylab='#samples')
	}
	
	conds= paste0(data[,5], ' # ',data[,6])
	condN6= condNumb(targetMarks= topMarks, conds= conds, marks=data[,7])
	condN5= condNumb(targetMarks= topMarks[1:5], conds= conds, marks=data[,7])
	condN4= condNumb(targetMarks= topMarks[1:4], conds= conds, marks=data[,7])
	condN= paste0('tot=', length(unique(conds)), ' top6=', length(condN6), 
		' top5=', length(condN5), ' top4=', length(condN4))
	barplot(sort(table(conds[conds %in% condN6]), decreasing=T), las=3, main= condN, ylab='#samples')

dev.off()
```
